A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-25 tex/sec4-pre-synthesis.tex
[ tex ] More proofs in S4

@@

\Josh{Throughout this sub-section, we quantify over any bidirectional type system~$(\Sigma, \Omega)$, context $\Gamma : \Cxt_\Sigma(\emptyset)$, raw term $\erase\Gamma \vdash_{\Sigma, \erase\Omega} t$, mode~$d$, and type $A : \Type_\Sigma(\emptyset)$.}

@@

\begin{figure}
  \centering
  \small
  \begin{tabular}{ r r l }
    & & is in (generalised) mode~$d$, \\
    $\smash{\boxed{V |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}\,v\,e}}}$
    & The raw term~$\isTerm{t}$\hspace{-.6em}
    & misses some type annotation iff $v = \False$, and \\
    & & has an outermost mode cast iff $e = \False$
  \end{tabular}
  \begin{mathpar}
    \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn\,\True\,\True}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,v\,e}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn\,v\,\True}}\,\SynRule{Anno}
    \and
    \inferrule{V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,v\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,\False\,\False}}\,\SynRule{Missing}
    \and
    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,v\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk\,v\,\False}}\,\ChkRule{Sub}
    \and
    \inferrule{V, \vec x_1 |-_{\Sigma, \Omega} \isTerm{t_1}^{\dir{d_1}\,v_1\,e_1} \\ \cdots \\ V, \vec x_n |-_{\Sigma, \Omega} \isTerm{t_n}^{\dir{d_n}\,v_n\,e_n}}
    {V |-_{\Sigma, \Omega} \tmOpts^{\dir{d}\,(\bigwedge_i v_i)\,\True}}\,\Rule{Op}
  \end{mathpar}
  \caption{Generalised mode derivations}
\end{figure}

\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-preprocessing}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
\end{theorem}

\begin{proof}
By induction on~$t$, we prove the rest of the theorem statement
\begin{equation}\label{eq:preprocess}
\forall d.~\exists v, e.~V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}
\end{equation}
in conjunction with
\begin{equation}\label{eq:preprocess'}
V |- t^{\syn\,\True\,\True} ~~\vee~~ V |- t^{\chk\,\True\,\True} ~~\vee~~ \exists e.~V |- t^{\chk\,\False\,e}
\end{equation}
Intuitively, statement~(\ref{eq:preprocess'}) gives a precise classification of a term as either a `real' synthesising term (i.e.~without missing annotations or an outermost mode cast), a `real' checking term, or a term that has missing annotations.
After a derivation is constructed for statement~(\ref{eq:preprocess'}), we can perform some slight adjustment to obtain a desired derivation for statement~(\ref{eq:preprocess}) depending on~$d$:
\begin{itemize}
\item For the first disjunct with $\dir{d} = {\syn}$, the second disjunct with $\dir{d} = {\chk}$, and the third disjunct with $\dir{d} = {\chk}$:
The mode we have matches the mode to be established, so no adjustment is needed.
\item For the first disjunct with $\dir{d} = {\chk}$:
Apply $\ChkRule{Sub}$.
\item For the second disjunct with $\dir{d} = {\syn}$:
Apply $\SynRule{Missing}$.
\item For the third disjunct with $\dir{d} = {\syn}$:
If $e = \False$, then the outermost rule of the derivation must be $\ChkRule{Sub}$, removing which we get a derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$;
otherwise $e = \True$, in which case we can apply $\SynRule{Missing}$.
\end{itemize}
Now we prove statement~(\ref{eq:preprocess'}) with a case analysis on~$t$.
\begin{itemize}
\item $\Rule{Var}$:
Establish the first disjunct using $\SynRule{Var}$.
\item $\Rule{Anno}$ ($t \bbcolon A$):
By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t^{\chk\,v\,e}$ for some boolean $v$~and~$e$.
If $v = \False$, establish the third conjunct using $\SynRule{Anno}$ and $\ChkRule{Sub}$; otherwise, establish the first conjunct using $\ChkRule{Sub}$.
\item $\Rule{Op}$:
By statement~(\ref{eq:preprocess}) in the induction hypothesis, there is a derivation of $V |-_{\Sigma,\Omega} t_i^{\dir{d_i}\,v_i\,e_i}$ for all the sub-terms~$t_i$.
\end{itemize}
\end{proof}

@@

@@

@@

\begin{proof}
By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
If $v = \True$, then $V |-_{\Sigma,\Omega} t^\dir{d}$ by \cref{thm:Pre?-true}; otherwise $V \not|-_{\Sigma,\Omega} t^\dir{d}$ by \cref{thm:Pre?-false}.
\end{proof}

\begin{corollary}
If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,e}$ for some boolean~$e$.
\end{corollary}

\begin{corollary}
If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,e}$ for some boolean~$e$.
\end{corollary}

\begin{proof}
By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,v\,e}$ for some boolean $v$~and~$e$.
Either $v = \True$ or $v = \False$.
If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $v$ has to be $\True$ since $v = \False$ leads to a contradiction with \cref{thm:Pre?-false}.
Symmetrically, if $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $v$ has to be $\False$ so as not to contradict \cref{thm:Pre?-true}.
\end{proof}

>>>>

\Josh{Overview of the section}

@@

%\Josh{Throughout this sub-section, we quantify over any bidirectional type system~$(\Sigma, \Omega)$, context $\Gamma : \Cxt_\Sigma(\emptyset)$, raw term $\erase\Gamma \vdash_{\Sigma, \erase\Omega} t$, mode~$d$, and type $A : \Type_\Sigma(\emptyset)$.}

@@

\begin{figure}
  \centering
  \small
  \begin{tabular}{ r r l }
    & & is in (generalised) mode~$d$, \\
    $\smash{\boxed{V |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}\,v\,e}}}$
    & The raw term~$\isTerm{t}$\hspace{-.6em}
    & misses some type annotation iff $v = \False$, and \\
    & & is in (generalised) mode~$d$ due to an outermost mode cast iff $e = \False$
  \end{tabular}
  \begin{mathpar}
    \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn\,\True\,\True}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,v\,e}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn\,v\,\True}}\,\SynRule{Anno}
    \and
    \inferrule{V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,v\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,\False\,\False}}\,\SynRule{Missing}
    \and
    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,v\,\True}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk\,v\,\False}}\,\ChkRule{Sub}
    \and
    \inferrule{V, \vec x_1 |-_{\Sigma, \Omega} \isTerm{t_1}^{\dir{d_1}\,v_1\,e_1} \\ \cdots \\ V, \vec x_n |-_{\Sigma, \Omega} \isTerm{t_n}^{\dir{d_n}\,v_n\,e_n}}
    {V |-_{\Sigma, \Omega} \tmOpts^{\dir{d}\,(\bigwedge_i v_i)\,\True}}\,\Rule{Op}
  \end{mathpar}
  \caption{Generalised mode derivations}
\end{figure}

\begin{lemma}\label{thm:adjustment}
For any raw term $V |-_{\Sigma,\erase\Omega} t$, if
\begin{equation}\label{eq:classification}
V |- t^{\syn\,\True\,\True} ~~\vee~~ V |- t^{\chk\,\True\,\True} ~~\vee~~ \exists e.~V |- t^{\chk\,\False\,e}
\end{equation}
then for any mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some booleans $v$~and~$e$.
\end{lemma}

\begin{proof}
Case analysis on statement~(\ref{eq:classification}) and~$\dir{d}$.
%\begin{itemize}
%\item For the first disjunct with $\dir{d} = {\syn}$, the second disjunct with $\dir{d} = {\chk}$, and the third disjunct with $\dir{d} = {\chk}$:
%The mode we have matches the mode to be established, so no adjustment is needed.
%\item For the first disjunct with $\dir{d} = {\chk}$:
%Apply $\ChkRule{Sub}$.
%\item For the second disjunct with $\dir{d} = {\syn}$:
%Apply $\SynRule{Missing}$.
%\item For the third disjunct with $\dir{d} = {\syn}$:
%If $e = \False$, then the outermost rule of the derivation must be $\ChkRule{Sub}$, removing which we get a derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$;
%otherwise $e = \True$, in which case we can apply $\SynRule{Missing}$.
%\vspace{-\topsep-\baselineskip}
%\end{itemize}
\end{proof}

Intuitively, statement~(\ref{eq:classification}) gives a precise classification of a term as either a `real' synthesising term (i.e.~without missing annotations or an outermost mode cast), a real checking term, or a term that has missing annotations.
The lemma performs some slight adjustment that involves only outermost mode casts to switch an input derivation to any required mode.

\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-preprocessing}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some booleans $v$~and~$e$.
\end{theorem}

\begin{proof}
It suffices to prove statement~(\ref{eq:classification}) by induction on~$t$ and then apply \cref{thm:adjustment}.
\begin{itemize}
\item $\Rule{Var}$:
A variable is a real synthesising term, so we establish the first disjunct using $\SynRule{Var}$.
\item $\Rule{Anno}$:
Let $t = (t' \bbcolon A)$.
By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,v\,e}$ for some booleans $v$~and~$e$.
If $v = \False$, then $t'$, and therefore~$t$, have missing annotations, and we establish the third disjunct using $\SynRule{Anno}$ and $\ChkRule{Sub}$; otherwise, $t$~is a real synthesising term, and we establish the first disjunct using $\ChkRule{Sub}$.
\item $\Rule{Op}$:
Let $t = \tmOpts$.
By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,v_i\,e_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ and we get a derivation of $V |-_{\Sigma,\Omega} t^{\dir d\,v\,\True}$ where $v = \bigwedge_i v_i$.
If $v = \True$, then $t$~is a real synthesising or checking term, and we establish the first or second disjunct depending on~$\dir d$; otherwise, some~$t_i$, and therefore~$t$, have missing annotations, and we establish the third disjunct, applying $\ChkRule{Sub}$ to get the mode right if $\dir d = {\syn}$.
\vspace{-\topsep-\baselineskip}
\end{itemize}
\end{proof}

\Josh{Separating generalised mode preprocessing into a lemma and a theorem to simplify the case analyses}

@@

\begin{proof}
Induction on the given derivation.
The $\SynRule{Missing}$ rule cannot appear, and the other rules are mapped to their counterparts.
\end{proof}

@@

\begin{proof}
Induction on the generalised mode derivation, in each case analysing an arbitrary mode derivation and showing that it cannot exist.
The key case is $\SynRule{Missing}$, where we have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\chk\,v\,\True}$ for some boolean~$v$.
We do not have an induction hypothesis and seem to get stuck, because $v$~is not necessarily~$\False$ --- indeed, $t$~may well be a real checking term, and this is where it is required to (but cannot) be synthesising.
But all that matters here is that $t$~is \emph{exactly} in checking mode: if we continue to analyse the sub-derivation, the outermost rule must be $\Rule{Op}$ with $\dir d = {\chk}$, implying that $t$~has to be an operation in checking mode.
Then a case analysis shows that it is impossible to have a (synthesising) mode derivation of $V |-_{\Sigma,\Omega} t^{\syn}$.
%\begin{itemize}
%\item $\SynRule{Anno}$:
%Any mode derivation must also use $\SynRule{Anno}$ as the outermost rule, and then the induction hypothesis suffices.
%\item $\ChkRule{Sub}$:
%We have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\False\,\True}$.
%Any mode derivation of $V |-_{\Sigma,\Omega} t^{\chk}$ must use $\ChkRule{Chk}$ as the outermost rule.
%(The $\Rule{Op}$ rule with $\dir d = {\chk}$ is impossible because the sub-derivation shows that $t$~is exactly a synthesising term.)
%We then have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn}$, and can use the induction hypothesis.
%\end{itemize}
\end{proof}

@@

\begin{proof}
By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,v\,e}$ for some booleans $v$~and~$e$, and then simply check~$v$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
\end{proof}

\begin{corollary}\label{thm:toPre?-true}
If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\True\,e}$ for some boolean~$e$.
\end{corollary}

\begin{corollary}\label{thm:toPre?-false}
If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\False\,e}$ for some boolean~$e$.
\end{corollary}

Whereas it is possible to prove \cref{thm:toPre?-true} directly by induction, converting a mode derivation to a generalised one, the negated antecedent of \cref{thm:toPre?-false} does not provide useful information.
To prove \cref{thm:toPre?-false}, basically we have to construct a generalised mode derivation from scratch for a raw term, but this is exactly generalised mode preprocessing, so we can simply reuse \cref{thm:generalised-mode-preprocessing}.
And \cref{thm:toPre?-true} can also be proved easily in the same way.

\begin{proof}[Proofs of \cref{thm:toPre?-true,thm:toPre?-false}]

By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,v\,e}$ for some booleans $v$~and~$e$.
If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $v$ has to be $\True$ since $v = \False$ leads to a contradiction with \cref{thm:Pre?-false}.
Symmetrically, if $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $v$ has to be $\False$ so as not to contradict \cref{thm:Pre?-true}.
\end{proof}
